---
output: github_document
always_allow_html: yes
---

<a href="https://zenodo.org/badge/latestdoi/268765075"><img src="https://zenodo.org/badge/268765075.svg" alt="DOI"></a>

<!-- README.md is generated from README.Rmd. Please edit that file -->
<!-- devtools::build_readme() -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# MiscMetabar

The goal of MiscMetabar is to complete the great packages [dada2](https://benjjneb.github.io/dada2/index.html), [phyloseq](https://joey711.github.io/phyloseq/) and [targets](https://books.ropensci.org/targets/). 
See the pkdown site [here](https://adrientaudiere.github.io/MiscMetabar/).

## Installation

There is no CRAN or bioconductor version of MiscMetabar for now (work in progress).

You can install the stable version from [GitHub](https://github.com/) with:

```{r, message=FALSE}
install.packages("devtools")
devtools::install_github("adrientaudiere/MiscMetabar")
```

You can install the developement version from [GitHub](https://github.com/) with:

```{r, message=FALSE}
install.packages("devtools")
devtools::install_github("adrientaudiere/MiscMetabar", ref = "dev")
```

## Some use of MiscMetabar

### Summarize a physeq object

```{r example, message=FALSE}
library("MiscMetabar")
library("phyloseq")
library("magrittr")
data("data_fungi")
summary_plot_pq(data_fungi)
```

### Circle for visualize distribution of taxa in function of samples variables

```{r}
circle_pq(data_fungi, "Height", taxa = "Class", add_nb_seq = F)
```
