---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# MiscMetabar

The goal of MiscMetabar is to complete the great package dada2 and phyloseq. 

## Installation

There is no CRAN version of MiscMetabar for now.

You can install the development version from [GitHub](https://github.com/) with:

```{r installation}
# install.packages("devtools")
devtools::install_github("adrientaudiere/MiscMetabar")
```

## Example

This is a basic example which shows you how to solve a common problem:

### Summarize a physeq object

```{r example}
library("MiscMetabar")
library("phyloseq")
data("esophagus")
summary_plot_phyloseq(esophagus)
```

### Create an interactive table of the tax_table

```{r}
data("GlobalPatterns")
tax_datatable(subset_taxa(GlobalPatterns,
             rowSums(GlobalPatterns@otu_table)>100000))
```

### Sankey diagramme of the tax_table

```{r}
GP <- subset_taxa(GlobalPatterns, GlobalPatterns@tax_table[,1] == "Archaea")
sankey_phyloseq(GP, taxa = c(1:5))
```

### Circle for visualise distribution of taxa in function of samples variables

```{r}
otu_circle(GP, 'SampleType', taxa = "Class")
```
