---
output: github_document
always_allow_html: yes
---

![R](https://img.shields.io/badge/r-%23276DC3.svg?style=for-the-badge&logo=r&logoColor=white)
<a href="https://zenodo.org/badge/latestdoi/268765075"><img src="https://zenodo.org/badge/268765075.svg" alt="DOI"></a>
[![codecov](https://codecov.io/gh/adrientaudiere/MiscMetabar/graph/badge.svg?token=NXFRSIKYC0)](https://codecov.io/gh/adrientaudiere/MiscMetabar)
[![Contributor Covenant](https://img.shields.io/badge/Contributor%20Covenant-2.1-4baaaa.svg)](code_of_conduct.md) 
[![License: GPL v3](https://img.shields.io/badge/License-GPL%20v3-blue.svg)](http://www.gnu.org/licenses/gpl-3.0)

<!-- README.md is generated from README.Rmd. Please edit that file -->
<!-- devtools::build_readme() -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# MiscMetabar

The goal of MiscMetabar is to complete the great packages [dada2](https://benjjneb.github.io/dada2/index.html), [phyloseq](https://joey711.github.io/phyloseq/) and [targets](https://books.ropensci.org/targets/). 
See the pkdown site [here](https://adrientaudiere.github.io/MiscMetabar/).

## Installation

There is no CRAN or bioconductor version of MiscMetabar for now (work in progress).

You can install the stable version from [GitHub](https://github.com/) with:

```{r, message=FALSE, eval=FALSE}
install.packages("devtools")
devtools::install_github("adrientaudiere/MiscMetabar")
```

You can install the developement version from [GitHub](https://github.com/) with:

```{r, message=FALSE, eval=FALSE}
install.packages("devtools")
devtools::install_github("adrientaudiere/MiscMetabar", ref = "dev")
```

## Some use of MiscMetabar

See vignettes in the [MiscMetabar](https://adrientaudiere.github.io/MiscMetabar/) website for more examples.


### Summarize a physeq object

```{r example, message=FALSE}
library("MiscMetabar")
library("phyloseq")
library("magrittr")
data("data_fungi")
summary_plot_pq(data_fungi)
```

### Alpha-diversity analysis

```{r, fig.cap="Hill number 1"}
p <- MiscMetabar::hill_pq(data_fungi, variable = "Height")
p$plot_Hill_0
```

```{r, fig.cap="Result of the Tuckey post-hoc test"}
p$plot_tuckey
```

### Beta-diversity analysis

```{r}
ggvenn_pq(data_fungi, fact = "Height") +
  ggplot2::scale_fill_distiller(palette = "BuPu", direction = 1) +
  labs(title = "Share number of ASV among Height in tree")
```

### Installation of other softwares for debian Linux distributions

#### blastn

```sh
sudo apt-get install ncbi-blast+
```

#### vsearch

```sh
sudo apt-get install vsearch
```
