---
output: github_document
always_allow_html: yes
---

<a href="https://zenodo.org/badge/latestdoi/268765075"><img src="https://zenodo.org/badge/268765075.svg" alt="DOI"></a>

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# MiscMetabar

The goal of MiscMetabar is to complete the great packages [dada2](https://benjjneb.github.io/dada2/index.html), [phyloseq](https://joey711.github.io/phyloseq/) and [targets](https://books.ropensci.org/targets/). 
See the pkdown site [here](https://adrientaudiere.github.io/MiscMetabar/).

## Installation

There is no CRAN or bioconductor version of MiscMetabar for now (work in progress).

You can install the stable version from [GitHub](https://github.com/) with:

```{r installation, message=FALSE}
install.packages("devtools")
devtools::install_github("adrientaudiere/MiscMetabar")
```

You can install the developement version from [GitHub](https://github.com/) with:

```{r installation, message=FALSE}
install.packages("devtools")
devtools::install_github("adrientaudiere/MiscMetabar", ref="dev")
```

## Some use of MiscMetabar

### Summarize a physeq object

```{r example, message=FALSE}
library("MiscMetabar")
library("phyloseq")
library("magrittr")
data("data_fungi")
summary_plot_phyloseq(data_fungi)
```

### Create an interactive table of the tax_table

```{r}
data("GlobalPatterns")
tax_datatable(subset_taxa(
  GlobalPatterns,
  rowSums(GlobalPatterns@otu_table) > 100000
))
```

### Sankey diagramme of the tax_table

```{r}
gp <- subset_taxa(GlobalPatterns, GlobalPatterns@tax_table[, 1] == "Archaea")
sankey_phyloseq(gp, taxa = c(1:5))
```

### Circle for visualize distribution of taxa in function of samples variables

```{r}
otu_circle(gp, "SampleType", taxa = "Class")
```
