---
title: "alpha-div"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{alpha-div}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MiscMetabar)
```
### Alpha diversity

Voir aussi le [tutoriel](https://microbiome.github.io/tutorials/Alphadiversity.html) du package microbiome. 


#### Hill number

Numerous metrics of diversity exist. 
Hill numbers [^Hill73] is a kind of general framework for alpha diversity index. 

[^Hill73]: Hill MO. 1973. Diversity and evenness: a unifying notation and its consequences. Ecology 54, 427-473.

```r
vegan::renyi()
```

#### Test for difference in diversity (hill number)

One way to keep into account for difference in the number of sequences per samples is to use a Tukey test on a linear model with the square roots of the number of sequence as the first explanatory variable of the linear model [^Balint15].

[^Balint15]: Bálint M et al. 2015. Relocation, high-latitude warming and host genetic identity shape the foliar fungal microbiome of poplars. Molecular Ecology 24, 235-248. https://doi.org/10.1111/mec.13018

```r
stats::TukeyHSD()
MiscMetabar::hill_tuckey_phyloseq()
```

#### Testing differences in alpha diversity following the [tutorial](https://microbiome.github.io/tutorials/Alphadiversity.html) of the microbiome package. 

We recommend the non-parametric [Kolmogorov-Smirnov test](https://www.rdocumentation.org/packages/dgof/versions/1.2/topics/ks.test) for two-group comparisons when there are no relevant covariates.

```r
# Construct the data
d <- meta(pseq)
d$diversity <- microbiome::diversity(pseq, "shannon")$shannon
# Split the values by group
spl <- split(d$diversity, d$sex)
# Kolmogorov-Smironv test
pv <- ks.test(spl$female, spl$male)$p.value
# Adjust the p-value
padj <- p.adjust(pv)
