---
title: "alpha-div"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{alpha-div}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MiscMetabar)
data(data_fungi)
```

### Alpha diversity

Voir aussi le [tutoriel](https://microbiome.github.io/tutorials/Alphadiversity.html) du package microbiome. 


#### Hill number

Numerous metrics of diversity exist. 
Hill numbers [^Hill73] is a kind of general framework for alpha diversity index. 

[^Hill73]: Hill MO. 1973. Diversity and evenness: a unifying notation and its consequences. Ecology 54, 427-473.

```{r}
renyi_res <- vegan::renyi(data_fungi@otu_table)
head(renyi_res)
```

#### Test for difference in diversity (hill number)

One way to keep into account for difference in the number of sequences per samples is to use a Tukey test on a linear model with the square roots of the number of sequence as the first explanatory variable of the linear model [^Balint15].

[^Balint15]: Bálint M et al. 2015. Relocation, high-latitude warming and host genetic identity shape the foliar fungal microbiome of poplars. Molecular Ecology 24, 235-248. https://doi.org/10.1111/mec.13018

```{r, fig.cap="Hill number 1"}
p <- MiscMetabar::hill_pq(data_fungi, variable = "Height")
p$plot_Hill_0
```

```{r, fig.cap="Result of the Tuckey post-hoc test"}
p$plot_tuckey
```

See also the [tutorial](https://microbiome.github.io/tutorials/Alphadiversity.html) of the microbiome package for an alternative using the non-parametric [Kolmogorov-Smirnov test](https://www.rdocumentation.org/packages/dgof/versions/1.2/topics/ks.test) for two-group comparisons when there are no relevant covariates.