---
title: "beta-div"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{beta-div}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

```{r setup, message=FALSE}
library(MiscMetabar)
data(data_fungi)
```

## Permanova

```{r}
data_fungi_woNA4height <- subset_samples(data_fungi, !is.na(data_fungi@sam_data$Height))
res_ado <- adonis_pq(data_fungi_woNA4height, "Tree_name+Height")
knitr::kable(res_ado)
```


## Graph Test

```{r}
data_fungi_woNA4height <- subset_samples(data_fungi, !is.na(data_fungi@sam_data$Height))
graph_test_pq(data_fungi_woNA4height, "Height")
```

## Circle of ASVs

```{r}
circle_pq(data_fungi_woNA4height, "Height")
```

## Compare two (group of) samples

### Biplot


```{r}
data_fungi_low_high <- subset_samples(
  data_fungi,
  data_fungi@sam_data$Height %in%
    c("Low", "High")
)
data_fungi_low_high <- subset_taxa_pq(
  data_fungi_low_high,
  taxa_sums(data_fungi_low_high) > 5000
)
biplot_pq(data_fungi_low_high, fact = "Height", merge_sample_by = "Height")
```

### Compare two (group of) samples with a table

```{r}
compare_pairs_pq(data_fungi_low_high,
  bifactor = "Height",
  merge_sample_by = "Height", modality = "Time"
)
```

## Venn diagram

```{r}
library("grid")
venn_pq(data_fungi, fact = "Height")
```

```{r}
ggvenn_pq(data_fungi, fact = "Height") +
  ggplot2::scale_fill_distiller(palette = "BuPu", direction = 1) +
  labs(title = "Share number of ASV among Height in tree")
```

```{r}
ggvenn_pq(data_fungi, fact = "Height", min_nb_seq = 5000) +
  ggplot2::scale_fill_distiller(palette = "BuPu", direction = 1) +
  labs(title = "Share number of ASV with more than 5000 seqs")
```

```{r}
ggvenn_pq(data_fungi,
  fact = "Height", taxonomic_rank = "Genus",
  min_nb_seq = 100
) +
  ggplot2::scale_fill_distiller(palette = "BuPu", direction = 1) +
  labs(title = "Share number of Genus represented by at least one ASV with more than 100 seqs")
```

## Upset plot

Venn diagram can quickly become complex to read when the number of modalities increase. One graphical solution is upset plot. MiscMetabar propose a solution based on the package [ComplexUpset](https://krassowski.github.io/complex-upset/).
```{r}
upset_pq(data_fungi, fact = "Height")
```

```{r}
upset_pq(data_fungi, fact = "Time")
```

`ComplexUpset` package allow powerful configuration of you plot as you can see in the following figure.

```{r}
upset_pq(
  data_fungi,
  fact = "Time",
  width_ratio = 0.2,
  annotations = list(
    "Sequences per ASV \n (log10)" = (
      ggplot(mapping = aes(y = log10(Abundance)))
      +
        geom_jitter(aes(
          color =
            Abundance
        ), na.rm = TRUE)
        +
        geom_violin(alpha = 0.5, na.rm = TRUE) +
        theme(legend.key.size = unit(0.2, "cm")) +
        theme(axis.text = element_text(size = 12))
    ),
    "ASV per phylum" = (
      ggplot(mapping = aes(fill = Phylum))
      +
        geom_bar() +
        ylab("ASV per phylum") +
        theme(legend.key.size = unit(0.2, "cm")) +
        theme(axis.text = element_text(size = 12))
    )
  )
)
```

## Change in abundance across a factor 

### Benchdamic

There is a lot of available methods. Please refer to R package [benchdamic](https://github.com/mcalgaro93/benchdamic) for a list of method and a implementation of a benchmark for your data. 

### Library requirement for Debian Linux OS

```sh
sudo apt-get install libgsl-dev libmpfr-dev
```

### Using Deseq2 package

```{r}
data("GlobalPatterns")
GP <- subset_samples(
  GlobalPatterns,
  GlobalPatterns@sam_data$SampleType %in% c("Soil", "Skin")
)

plot_deseq2_pq(GP, c("SampleType", "Soil", "Skin"), alpha = 0.001)
```

