---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction to MiscMetabar : a R packages to facilitate visualization and reproductibility in metabarcoding analysis

## Raison d'être

- Complete R packages dada2 and phyloseq
- Useful visualizations (`biplot_pq`, `circle_pq`) 
- Facilitate the use of targets 

## Quick overview

### Summarize a physeq object

```{r example, message=FALSE}
library("MiscMetabar")
library("phyloseq")
library("magrittr")
data("data_fungi")
summary_plot_pq(data_fungi)
```


### Create an interactive table of the tax_table

```{r results="asis"}
data("GlobalPatterns")
tax_datatable(subset_taxa(
  GlobalPatterns,
  rowSums(GlobalPatterns@otu_table) > 100000
))
```

### Sankey diagramme of the tax_table

```{r}
gp <- subset_taxa(GlobalPatterns, GlobalPatterns@tax_table[, 1] == "Archaea")
sankey_pq(gp, taxa = c(1:5))
```


### Circle for visualize distribution of taxa in function of samples variables

```{r}
circle_pq(gp, "SampleType", taxa = "Class")
```