---
title: "Accumulation and rarefaction curves (WORK IN PROGRESS)"
output: rmarkdown::html_vignette
bibliography: ../bibliography.bib
orphan: true
vignette: >
  %\VignetteIndexEntry{Accumulation and rarefaction curves}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---


/!\ (WORK IN PROGRESS) /!\



```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

Accumulation and rarefaction curves are fundamental tools in biodiversity
analysis. They address a key challenge: **comparing diversity across
communities that differ in sampling effort**. In metabarcoding, this
challenge is compounded by a second, more fundamental one: the relationship
between sequence counts and the actual biological diversity in the field.

This vignette explains the conceptual choices involved and illustrates how
to compute and visualise these curves using `hill_acc_pq()` and
`profile_hill_pq()`, two MiscMetabar functions built on the
[divent](https://ericmarcon.github.io/divent/) package [@Marcon2015].

## Which function should I use?

```{r decision-tree, echo = FALSE, fig.cap = "Decision tree for choosing the appropriate diversity curve in metabarcoding."}
DiagrammeR::grViz('
digraph decision_tree {

  graph [rankdir = LR, fontname = Helvetica, fontsize = 11,
         bgcolor = "#FAFAFA", pad = 0.4]
  node  [fontname = Helvetica, fontsize = 10, margin = "0.18,0.10",
         penwidth = 1.2]
  edge  [fontname = Helvetica, fontsize = 9, penwidth = 1.1]

  ROOT [label = "Biological\nquestion?",
        shape = diamond, style = filled,
        fillcolor = "#AED6F1", color = "#2980B9"]

  Q1 [label = "Is sequencing depth\nsufficient?",
      shape = rect, style = "filled,rounded",
      fillcolor = "#D6EAF8", color = "#2980B9"]

  Q2 [label = "Does spatial / temporal\nsampling cover diversity?",
      shape = rect, style = "filled,rounded",
      fillcolor = "#D6EAF8", color = "#2980B9"]

  Q3 [label = "What is the community\nevenness structure?",
      shape = rect, style = "filled,rounded",
      fillcolor = "#D6EAF8", color = "#2980B9"]

  Q4 [label = "Are samples comparable\nin detection coverage?",
      shape = rect, style = "filled,rounded",
      fillcolor = "#D6EAF8", color = "#2980B9"]

  T2 [label = "Trust read counts\nas abundance proxies?",
      shape = diamond, style = filled,
      fillcolor = "#FDEBD0", color = "#E67E22"]

  T3 [label = "Trust read counts\nas abundance proxies?",
      shape = diamond, style = filled,
      fillcolor = "#FDEBD0", color = "#E67E22"]

  F1 [label = "hill_acc_pq(\n  type = individual\n)\nx-axis: no. of sequences",
      shape = rect, style = "filled,rounded",
      fillcolor = "#A9DFBF", color = "#1E8449"]

  F2a [label = "hill_acc_pq(\n  type = sample\n)\nx-axis: no. of samples",
       shape = rect, style = "filled,rounded",
       fillcolor = "#A9DFBF", color = "#1E8449"]

  F2b [label = "as_binary_otu_table() +\nhill_acc_pq(\n  type = sample\n)\nx-axis: no. of samples",
       shape = rect, style = "filled,rounded",
       fillcolor = "#D5F5E3", color = "#1E8449"]

  F3a [label = "profile_hill_pq()\nx-axis: Hill order q",
       shape = rect, style = "filled,rounded",
       fillcolor = "#A9DFBF", color = "#1E8449"]

  F3b [label = "as_binary_otu_table() +\nprofile_hill_pq()\nx-axis: Hill order q",
       shape = rect, style = "filled,rounded",
       fillcolor = "#D5F5E3", color = "#1E8449"]

  F4 [label = "iNEXT.4steps\n(see section below)\nx-axis: sample coverage",
      shape = rect, style = "filled,rounded",
      fillcolor = "#FAD7A0", color = "#B7770D"]

  ROOT -> Q1
  ROOT -> Q2
  ROOT -> Q3
  ROOT -> Q4

  Q1  -> F1
  Q2  -> T2
  Q3  -> T3
  Q4  -> F4

  T2  -> F2a [label = "yes"]
  T2  -> F2b [label = "no / uncertain"]
  T3  -> F3a [label = "yes"]
  T3  -> F3b [label = "no / uncertain"]
}
')
```

Read the sections below for the conceptual background behind each choice.
The short answer on **trusting read counts**: use abundance-based approaches
when your system is well-controlled (mock communities, microbiomes, calibrated
protocols); prefer incidence-based when PCR biases or cross-contamination are
likely — as is common in environmental surveys.

## The eDNA challenge: sequences are not individuals

In traditional ecology, rarefaction standardises samples by drawing a fixed
number of **individuals**, each of which unambiguously represents one
organism. In metabarcoding, the recorded entities are **DNA sequences**, not
individuals. This creates a fundamental ambiguity: one species can be
represented by tens of thousands of reads (high DNA concentration, efficient
primer binding) or by a handful (low biomass, poor amplification) regardless
of its true ecological abundance [@Alberdi2019; @Calderon2019]. Controlled
mock-community experiments confirm this: read counts per species vary over an
order of magnitude even at equal molar input, and correlations with true
abundance hold only for dominant taxa [@Skelton2023; @Shelton2023].

This weakens the link between sequence counts and actual species abundance,
and has practical consequences for how we measure diversity.

## Abundance-based vs. incidence-based approaches

Two broad strategies exist for weighting the contribution of each OTU to
diversity [@Alberdi2019]:

- **Abundance-based approach**: uses the number of DNA sequences assigned to
  each OTU as a proxy for its relative abundance. This is the default
  approach in most metabarcoding workflows and is appropriate when sequence
  counts are a reasonable proxy for relative biomass (e.g. in gut microbiome
  studies where microbial cells are well-represented in proportion to their
  abundance).

- **Incidence-based approach**: ignores within-sample sequence counts and
  instead uses the number of **samples** in which each OTU is detected
  (presence/absence). Because it discards quantitative information, it is
  less sensitive to amplification biases, cross-sample contamination, or
  tag-jump artefacts — all of which inflate the counts of rare sequences.
  Incidence-based Hill numbers are only meaningful at the level of the whole
  system (pool of samples), not for individual samples.

Both approaches converge at **q = 0** (species richness), because richness
almost ignores relative abundances. They diverge for **q > 0**, where the
abundance-based approach measures the effective number of equally *abundant*
OTUs and the incidence-based approach measures the effective number of
equally *frequent* (across samples) OTUs.

In practice, the choice depends on the research question and the degree of
trust in sequence counts. The two are complementary: computing both and
comparing them reveals how much the conclusions depend on quantitative
information.

## Hill numbers: a unified framework

Hill numbers [@Hill73] provide a family of diversity indices parameterised
by an order *q*:

- **q = 0**: species richness, treating all OTUs equally regardless of
  abundance. Highly sensitive to rare OTUs, which in metabarcoding are
  disproportionately likely to be sequencing artefacts.
- **q = 1**: exponential of Shannon entropy, weighting OTUs proportionally
  to their frequency. A balanced measure.
- **q = 2**: inverse Simpson concentration, dominated by common OTUs.
  Robust to rare artefactual sequences.

An important consequence for eDNA data: **higher q values are more robust to
data curation choices** [@Calderon2019; @Shirazi2021]. Rarefaction curves at
q = {1, 2} typically saturate well, while curves at q = 0 rarely saturate and
are heavily affected by data curation decisions (chimera removal, clustering
thresholds, etc.) and PCR variability across replicates. This makes q = 1 a
pragmatic default — it weights OTUs by their frequency without
disproportionately favouring rare or abundant ones.

For a thorough theoretical treatment of Hill Number, see
[Marcon (2018)](https://ericmarcon.github.io/MesuresBioDiv2/chap-Accumulation.html).

## Setup

```{r setup}
library(MiscMetabar)
library(phyloseq)
library(ggplot2)
library(patchwork)
set.seed(42)
data(data_fungi_mini)
```

We rarefy `data_fungi_mini` to an even sequencing depth of 500 sequences per
sample. This standardises sampling effort and is required by
individual-based rarefaction. In practice, rarefy to a higher depth if
possible. Recent analyses confirm that rarefaction remains the most robust
approach for controlling uneven sequencing depth in amplicon data
[@Schloss2024a; @Schloss2024b].

```{r}
dfm_rarefied <- rarefy_even_depth(
  data_fungi_mini,
  rngseed = 1,
  sample.size = 500
) 
dfm_rarefied <-  prune_samples(sample_names(dfm_rarefied) %in% sample(sample_names(dfm_rarefied), 25), dfm_rarefied)
dfm_rarefied
```

## Individual-based rarefaction (abundance-based)

Individual-based rarefaction plots Hill diversity as a function of the number
of sequences drawn. It answers: *is sequencing depth sufficient to
characterise diversity?* The function `hill_acc_pq()` with
`type = "individual"` wraps `divent::accum_hill()`.

Note that this is an **abundance-based** approach: the curve reflects the
sequence count distribution within each sample.

### Metacommunity level

Merge all samples into a single community to obtain a global rarefaction
curve:

```{r}
sample_data(dfm_rarefied)$metacommunity <- "all"
```

```{r, fig.cap="Individual-based rarefaction for the entire metacommunity (q = 1)"}
hill_acc_pq(
  dfm_rarefied,
  q = 1,
  type = "individual",
  merge_sample_by = "metacommunity",
  n_simulations = 10
) +
  no_legend() +
  ggtitle("Metacommunity rarefaction (q = 1)")
```

### Per-sample rarefaction

Examining individual samples reveals heterogeneity in diversity and sampling
completeness. Samples whose curve has not plateaued may harbour undetected
taxa.

```{r, fig.cap="Individual-based rarefaction for 5 random samples at different Hill orders"}
random_samples <- sample(sample_names(dfm_rarefied), 5)
dfm_5 <- prune_samples(random_samples, dfm_rarefied)

p_q0 <- hill_acc_pq(dfm_5, q = 0, type = "individual", n_simulations = 0) +
  ggtitle("q = 0 (richness)")
p_q1 <- hill_acc_pq(dfm_5, q = 1, type = "individual", n_simulations = 0) +
  ggtitle("q = 1 (Shannon)")
p_q2 <- hill_acc_pq(dfm_5, q = 2, type = "individual", n_simulations = 0) +
  ggtitle("q = 2 (Simpson)")

(p_q0 + p_q1 + p_q2) & no_legend()
```

At q = 0, curves rarely plateau because rare OTUs (many of which may be
artefacts) keep accumulating with additional sequences. At q = 1 and q = 2,
curves saturate faster because dominant OTUs are detected early. This
illustrates why q = {1, 2} provides more reliable sampling completeness
assessments [@Calderon2019].

### Rarefaction by habitat

Merging samples by the `Height` variable aggregates sequences within each
habitat, enabling comparison at the habitat scale:

```{r, fig.cap="Individual-based rarefaction by Height (q = 1)"}
hill_acc_pq(
  dfm_rarefied,
  q = 1,
  type = "individual",
  merge_sample_by = "Height",
  n_simulations = 0
) +
  ggtitle("Rarefaction by Height (q = 1)")
```

## Sample-based accumulation (abundance-based)

Sample-based accumulation plots Hill diversity as a function of the number
of **samples** added. It answers: *does spatial (or temporal) sampling
capture the full diversity?* It reflects turnover across samples
(beta-diversity), not sequencing depth within samples.

`hill_acc_pq()` with `type = "sample"` pools samples in random order across
`n_permutations` replicates and computes Hill diversity of the pooled
community at each step. The shaded band shows the confidence envelope.

Note: even though this approach aggregates samples, the pooled community
is still evaluated using **sequence counts** — it remains abundance-based.
Compare with the incidence-based approach in the next section.

```{r, fig.cap="Sample-based accumulation for the entire dataset (q = 1, 50 permutations)"}
hill_acc_pq(
  dfm_rarefied,
  q = 1,
  type = "sample",
  n_permutations = 50
) +
  ggtitle("Sample-based accumulation (q = 1)")
```

By splitting by habitat, one can assess whether diversity saturates within
each level and how much each level contributes to the total:

```{r, fig.cap="Sample-based accumulation by Height (q = 1)"}
hill_acc_pq(
  dfm_rarefied,
  q = 1,
  type = "sample",
  merge_sample_by = "Height",
  n_permutations = 50
) +
  ggtitle("Sample-based accumulation by Height (q = 1)")
```

## Incidence-based approach

When sequence counts are not trusted as proxies for abundance — for instance
due to primer binding biases or cross-contamination — an incidence-based
approach uses only presence/absence of OTUs across samples.
`as_binary_otu_table()` converts the OTU table to 0/1 before analysis.

Because incidence-based Hill numbers measure the effective number of OTUs
weighted by the fraction of samples they occur in, they are only meaningful
at the level of the entire system and are most naturally used with
`type = "sample"`.

```{r}
data(data_fungi)
dfm_binary <- as_binary_otu_table(data_fungi)
```

```{r, fig.cap="Incidence-based sample accumulation (q = 1)"}
hill_acc_pq(
  dfm_binary,
  q = 1,
  type = "sample",
  n_permutations = 50
) +
  ggtitle("Incidence-based sample accumulation (q = 1)")
```

```{r, fig.cap="Incidence-based sample accumulation by Height (q = 1)"}
data_fungi_woNA4Height <- subset_samples_pq(
  data_fungi,
  !is.na(data_fungi@sam_data$Height)
)
dfm_binary_height <- as_binary_otu_table(data_fungi_woNA4Height)

hill_acc_pq(
  dfm_binary_height,
  q = 1,
  type = "sample",
  merge_sample_by = "Height",
  n_permutations = 50
) +
  ggtitle("Incidence-based sample accumulation by Height (q = 1)")
```

Comparing the abundance-based and incidence-based accumulation curves reveals
how much the diversity ordering among habitats depends on quantitative
sequence information vs. simple detection patterns.

## Diversity profiles

Diversity profiles show how the Hill diversity index varies continuously
across orders *q*, for a fixed dataset. `profile_hill_pq()` wraps
`divent::profile_hill()`.

A steeply declining profile indicates an uneven community dominated by a few
OTUs. A flat profile indicates high evenness. The shape of the profile is
itself informative: two communities with identical diversity at q = 1 may
have very different profiles, revealing different abundance distributions.

### Abundance-based profiles

```{r, fig.cap="Abundance-based Hill diversity profiles per sample"}
profile_hill_pq(dfm_rarefied) + no_legend()
```

```{r, fig.cap="Abundance-based Hill diversity profiles merged by Height"}
profile_hill_pq(dfm_rarefied, merge_sample_by = "Height", n_simulations = 10)
```

### Incidence-based profiles

Applying `profile_hill_pq()` to a binary OTU table yields incidence-based
diversity profiles, where each OTU is weighted by the proportion of samples
it occurs in rather than its sequence count:

```{r, fig.cap="Incidence-based Hill diversity profiles merged by Height"}
profile_hill_pq(
  as_binary_otu_table(data_fungi_woNA4Height),
  merge_sample_by = "Height",
  n_simulations = 10
)
```

Contrasting the abundance-based and incidence-based profiles highlights which
diversity patterns are robust to the quantitative interpretation of sequence
counts and which depend on it.

## Beyond size-based rarefaction: coverage-based approaches

The approaches above standardise by **sample size** (number of sequences or
number of samples). An emerging alternative standardises by **sample
coverage** — the estimated fraction of the community already detected in a
sample [@Chao2020]. This matters because the same sequencing depth can yield
very different coverages depending on community evenness: a highly uneven
community (a few dominant OTUs) achieves high coverage quickly, while an
even one requires far more sequences.

Importantly, when the treatment or environment of interest also affects
community evenness (e.g. high land-use intensity simplifies communities),
size-based rarefaction conflates diversity loss with coverage gain.
@Kortmann2025 demonstrated this directly: coverage-standardised Hill numbers
revealed a 27–44% stronger diversity decline along a land-use gradient in
insects than size-based rarefaction suggested. The theoretical foundation
(sample completeness profiles parameterised by *q*) is formalised in
@Chao2020.

Coverage-based rarefaction is not yet implemented in `hill_acc_pq()`, but is
available via the [iNEXT.4steps](https://cran.r-project.org/package=iNEXT.4steps)
package, which accepts both abundance and incidence data and produces
rarefaction/extrapolation curves standardised by coverage for any Hill
order *q* [@Chao2020]. It is worth running alongside the curves produced
here whenever there is reason to suspect that sampling coverage varies across
the samples being compared.

## Summary

| Approach | Data | x-axis | Function | What it reveals |
|----------|------|--------|----------|-----------------|
| Individual-based rarefaction | Abundance | Number of sequences | `hill_acc_pq(type = "individual")` | Sequencing depth sufficiency |
| Sample-based accumulation | Abundance | Number of samples | `hill_acc_pq(type = "sample")` | Spatial turnover, beta-diversity |
| Incidence-based accumulation | Presence/absence | Number of samples | `as_binary_otu_table()` + `hill_acc_pq(type = "sample")` | Turnover robust to amplification biases |
| Abundance-based profiles | Abundance | Hill order *q* | `profile_hill_pq()` | OTU abundance distribution |
| Incidence-based profiles | Presence/absence | Hill order *q* | `as_binary_otu_table()` + `profile_hill_pq()` | OTU frequency distribution across samples |

## References

<div id="refs"></div>

## Session information

```{r}
sessionInfo()
```
