---
title: "Accumulation and rarefaction curves"
output: rmarkdown::html_vignette
bibliography: ../bibliography.bib
vignette: >
  %\VignetteIndexEntry{Accumulation and rarefaction curves}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

Accumulation and rarefaction curves are fundamental tools in biodiversity
analysis. They address a key challenge: **comparing diversity across
communities that differ in sampling effort**. This vignette covers the
theoretical background and illustrates how to compute and plot these
curves using MiscMetabar functions based on the
[divent](https://ericmarcon.github.io/divent/) package.

For a comprehensive theoretical treatment, see
[Marcon (2018)](https://ericmarcon.github.io/MesuresBioDiv2/chap-Accumulation.html).

## Rarefaction vs. accumulation

Two distinct types of curves exist:

- **Individual-based rarefaction** (or **size-based rarefaction**): plots
  diversity as a function of the **number of individuals** (in
  metabarcoding: the number of sequences). Given a fixed sample, it
  estimates the expected diversity if only a subset of individuals had
  been drawn. This allows fair comparison of communities sampled with
  different sequencing depths.

- **Sample-based accumulation** (or **species accumulation curves**, SAC):
  plots diversity as a function of the **number of samples** (e.g.,
  plots, sites, PCR replicates). It tracks how diversity grows as
  more sampling units are added.

## Hill numbers: a unified framework

Hill numbers [@Hill73] provide a family of diversity indices parameterized
by an order *q*:

- **q = 0**: species richness (number of species), sensitive to rare species.
- **q = 1**: the exponential of Shannon entropy, weighting species by their
  frequency equally.
- **q = 2**: the inverse Simpson concentration, dominated by common species.

By varying *q*, one obtains a **diversity profile** that fully
characterizes the diversity of a community. The divent package
[@Marcon2015] implements estimation and accumulation of Hill numbers with
proper bias correction.

## Setup

```{r setup}
library(MiscMetabar)
library(phyloseq)
library(ggplot2)
set.seed(42)
data(data_fungi_mini)
```

We rarefy `data_fungi_mini` to an even sequencing depth of 500 sequences per
sample, allowing faster computation of accumulation curves. In practice, you may want to rarefy to a higher depth or use the `hill_acc_pq()` function directly on the unrarefied data.

```{r}
dfm_rarefied <- rarefy_even_depth(
  data_fungi,
  rngseed = 1,
  sample.size = 500
)
dfm_rarefied
```

## Individual-based rarefaction (sequences as individuals)

Individual-based rarefaction curves show how estimated diversity grows with
the number of sequences drawn. The function `hill_acc_pq()` wraps
`divent::accum_hill()` and works directly on phyloseq objects.

### Metacommunity level

To compute the rarefaction curve for the entire dataset treated as a single
metacommunity, we merge all samples into one by adding a common grouping
variable. This represents the global diversity pool.

```{r}
# Add a dummy variable to merge all samples into a single metacommunity
sample_data(dfm_rarefied)$metacommunity <- "all"
```

```{r, fig.cap="Individual-based rarefaction for the entire metacommunity (q = 0, species richness)"}
p0 <-hill_acc_pq(
  dfm_rarefied,
  q = 0,
  merge_sample_by = "metacommunity",
  n_simulations = 0
) +
  no_legend() +
  ggtitle("Metacommunity rarefaction (q = 0)")
p0
```


### Five random samples

To examine per-sample rarefaction, we draw 5 samples at random. Each
curve shows how the diversity of a single sample grows with sequencing
depth.

```{r, fig.cap="Individual-based rarefaction for 5 random samples (q = 1)"}
random_samples <- sample(sample_names(dfm_rarefied), 5)
dfm_5 <- prune_samples(random_samples, dfm_rarefied)

hill_acc_pq(dfm_5, q = 1, n_simulations = 0) +
  ggtitle("Rarefaction curves for 5 random samples (q = 1)")
```

Samples that plateau early have lower diversity; samples whose curve is
still rising at the maximum depth may harbour additional undetected species.

```{r, fig.cap="Individual-based rarefaction for 5 random samples (q = 0, species richness)"}
p5_q0 <- hill_acc_pq(dfm_5, q = 0, n_simulations = 0) +
  ggtitle("Rarefaction curves for 5 random samples (q = 0)")
```

Changing the diversity order shifts the focus from rare to common species:

```{r, fig.cap="Individual-based rarefaction for 5 random samples at different Hill orders"}
p5_q1 <- hill_acc_pq(dfm_5, q = 1, n_simulations = 0) +
  ggtitle("Rarefaction curves for 5 random samples (q = 0)")
p5_q2 <- hill_acc_pq(dfm_5, q = 2, n_simulations = 0) +
  ggtitle("Rarefaction curves for 5 random samples (q = 0)")

library(patchwork)
p5_q0 + p5_q1 + p5_q2 & no_legend()
```

At q = 0, the curve keeps rising because rare species are still being
discovered. At higher orders, the curve saturates faster because common
species dominate the index.



### Samples merged by Height

Merging samples by the `Height` factor (Low, Middle, High) aggregates
sequences within each level, allowing comparison of diversity at the
habitat level.

```{r, fig.cap="Individual-based rarefaction by Height (q = 0)"}
hill_acc_pq(dfm_rarefied, q = 0, merge_sample_by = "Height", n_simulations = 0) +
  ggtitle("Rarefaction by Height (q = 0, richness)")
```


## Sample-based accumulation (number of samples)

Sample-based accumulation curves track how diversity grows as more
**sampling units** are added. This is fundamentally different from
individual-based rarefaction: it captures the spatial (or temporal if samples are time points)
turnover in species composition across samples.

The function `accu_plot()` wraps `vegan::specaccum()` and supports both
a global accumulation curve and per-factor curves.

### All dataset

```{r, fig.cap="Sample-based accumulation curve for the entire dataset with per-Height curves"}
accu_plot(dfm_rarefied, fact = "Height", add_nb_seq = FALSE)
```

The black curve ("all") shows the global accumulation. Per-Height curves
show how quickly diversity saturates within each habitat. If the
within-habitat curves plateau well below the global curve, this indicates
substantial beta-diversity (species turnover between habitats).

### By Height with sequence information

`accu_plot()` can also display the number of sequences on the x-axis
(`add_nb_seq = TRUE`), showing accumulation as a function of sequencing
effort.

```{r, fig.cap="Sample-based accumulation by Height with number of sequences on x-axis"}
accu_plot(dfm_rarefied, fact = "Height", add_nb_seq = TRUE, by.fact = TRUE) +
  theme(legend.position = "bottom")
```

```{r, fig.cap="Sample-based accumulation by Height (one line per sample, colored by Height)"}
accu_plot(dfm_rarefied, fact = "Height", add_nb_seq = TRUE) +
  theme(legend.position = "bottom")
```

### Balanced modality

When modalities have unequal sample sizes, `accu_plot_balanced_modality()`
subsamples to the smallest group size, providing a fairer comparison.

```{r, fig.cap="Balanced accumulation curves by Height"}
data(data_fungi)
data_fungi_woNA4Height <- subset_samples_pq(
  data_fungi,
  !is.na(data_fungi@sam_data$Height)
)
accu_plot_balanced_modality(data_fungi_woNA4Height, "Height", nperm = 3)
```

## Diversity profiles

Diversity profiles complement accumulation curves by showing how the
diversity index varies across Hill orders for a fixed sample size. The
function `profile_hill_pq()` wraps `divent::profile_hill()`.

```{r, fig.cap="Hill diversity profiles per sample"}
profile_hill_pq(dfm_rarefied) + no_legend()
```

```{r, fig.cap="Hill diversity profiles merged by Height"}
profile_hill_pq(dfm_rarefied, merge_sample_by = "Height", n_simulations = 10)
```

A steeply declining profile indicates an uneven community dominated by
a few species. A flat profile indicates high evenness.

```{r}
#TODO : interpretation
#profile_hill_pq(as_binary_otu_table(data_fungi), merge_sample_by = "Height",n_simulations=10)

```

## Summary

| Approach | x-axis | Function | What it reveals |
|----------|--------|----------|-----------------|
| Individual-based rarefaction | Number of sequences | `hill_acc_pq()` | Whether sequencing depth is sufficient |
| Sample-based accumulation | Number of samples | `accu_plot()` | Spatial turnover, beta-diversity |
| Diversity profile | Hill order *q* | `profile_hill_pq()` | Evenness structure |

## References

<div id="refs"></div>

## Session information

```{r}
sessionInfo()
```
