---
title: "Transformation and normalisation of phyloseq objects"
output: rmarkdown::html_vignette
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{Transformation and normalisation of phyloseq objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
```

```{r setup, message=FALSE}
library(MiscMetabar)
data(data_fungi_mini)
```

Choosing how to normalise or transform a count table is one of the most
consequential (and discussed) decisions in metabarcoding analysis.  Unequal
sequencing depths across samples introduce a systematic bias that can distort
every downstream step — alpha diversity, ordination, differential abundance
testing.  Yet every normalisation method makes its own assumptions and
trade-offs [@mcmurdie2014rarefaction].

This article surveys the methods available in **MiscMetabar**, shows their
effect on `data_fungi_mini`, and offers a decision guide to help you pick the
right approach.

---

## Why does it matter?

Amplicon sequencing returns *compositional* data: only the relative
proportions of taxa are observed, not their absolute abundances
[@gloor2017; @quinn2018].  On top of that, libraries differ in total read
count by one to two orders of magnitude, even within the same experiment.
Applying diversity or ordination methods to raw counts therefore conflates
biological signal with technical depth variation.

```{r depth-variation, fig.cap="Large variation in sequencing depth across samples in 'data_fungi_mini'."}
ggplot(data=tibble(x=sample_sums(data_fungi_mini)),aes(x=x)) + geom_histogram(color="black") + scale_x_log10() + labs(x="Number of sequences per samples")
```

---

## Overview of methods

| Family | Function | Key property |
|---|---|---|
| Presence/absence | `as_binary_otu_table()` | Ignores abundance entirely |
| Proportions (TSS) | `transform_pq(method="tss")` | Divides by library size |
| TSS + log | `normalize_prop_pq()` | TSS × constant, then log |
| Hellinger | `transform_pq(method="hellinger")` | Square-root of proportions |
| Centred log-ratio | `transform_pq(method="clr")` | Compositionally coherent |
| Robust CLR | `transform_pq(method="rclr")` | CLR robust to zeros |
| Log(1+x) | `transform_pq(method="log1p")` | Simple variance stabilisation |
| Z-score | `transform_pq(method="z")` | Per-taxon standardisation |
| Rank | `transform_pq(method="rank")` | Non-parametric, outlier robust |
| Rarefaction | `rarefy_pq()` | Subsampling to equal depth |
| SRS | `srs_pq()` | Rank-preserving subsampling [@heidrich2021] |
| GMPR | `gmpr_pq()` | Pairwise ratio geometric mean [@chen2018] |
| CSS | `css_pq()` | Cumulative sum scaling [@paulson2013] |
| TMM | `tmm_pq()` | Trimmed mean of M-values [@robinson2010] |
| VST | `vst_pq()` | Variance-stabilising (DESeq2) [@love2014] |
| Depth residuals | `mcknight_residuals_pq()` | Log-log regression residuals [@mcknight2019] |

---

## Exploring the transformations

### `transform_pq()` — a unified interface via `vegan::decostand`

`transform_pq()` wraps `vegan::decostand()` and handles the `taxa_are_rows`
orientation automatically.

```{r transform-pq, examplesIf = rlang::is_installed("vegan")}
data_tss <- transform_pq(data_fungi_mini, method = "tss")
data_hell <- transform_pq(data_fungi_mini, method = "hellinger")
data_clr <- transform_pq(data_fungi_mini, method = "clr")  # pseudocount = 1 by default
data_log1p <- transform_pq(data_fungi_mini, method = "log1p")

# Sample sums after TSS: all 1
round(range(sample_sums(data_tss)), 4)
```

**Hellinger** and **CLR** are particularly recommended for ordination
[@legendre2001; @gloor2017].  Hellinger distance on the transformed table
equals the Hellinger distance without transformation, which behaves better
than Bray-Curtis on sparse data.  CLR makes the data compositionally coherent
but requires replacing zeros (vegan uses a pseudocount internally).

### `normalize_prop_pq()` — TSS with log scaling (McKnight 2019)

Multiplies relative abundances by a constant (default 10 000) then applies a
log transformation.  This is the "rarefaction-free" normalisation proposed by
@mcknight2019.

```{r normalize-prop}
data_norm <- normalize_prop_pq(data_fungi_mini)
# All sample sums are equal after log transform
round(range(sample_sums(data_norm)), 2)
```

### `rarefy_pq()` — rarefaction with optional averaging

Rarefaction subsamples all libraries to the same depth, discarding reads.  It
remains the most widely used method despite debates about data loss
[@mcmurdie2014rarefaction; @mcknight2019].  `rarefy_pq()` allows averaging over `n`
random subsamplings to reduce stochasticity.

```{r rarefy}
data_rar1 <- rarefy_pq(data_fungi_mini, seed = 1)
data_rar10 <- rarefy_pq(data_fungi_mini, n = 10, seed = 1)

# Single rarefaction: all depths equal
unique(sample_sums(data_rar1))
# Averaged: sample sums close but not integer
round(range(sample_sums(data_rar10)), 1)
```

### `srs_pq()` — Scaling with Ranked Subsampling

SRS [@heidrich2021] scales to a common library size while **preserving the
rank order** of OTU abundances, avoiding the randomness of rarefaction and
the data loss associated with it.

```{r srs, eval = rlang::is_installed("SRS")}
data_srs <- srs_pq(data_fungi_mini)
round(range(sample_sums(data_srs)), 0)
```

### `gmpr_pq()` — Geometric Mean of Pairwise Ratios

GMPR [@chen2018] estimates size factors from the geometric mean of median
pairwise ratios between samples, making it robust to the high proportion of
zeros in microbial OTU tables.

```{r gmpr}
data_gmpr <- gmpr_pq(data_fungi_mini)
round(range(sample_sums(data_gmpr)), 0)
```

### `css_pq()` — Cumulative Sum Scaling

CSS [@paulson2013] normalises each sample by the cumulative sum of counts up
to a data-driven percentile, rather than the total.  It is less sensitive to
a few very abundant OTUs.

```{r css, eval = rlang::is_installed("metagenomeSeq")}
data_css <- css_pq(data_fungi_mini)
```

### `tmm_pq()` — Trimmed Mean of M-values

TMM [@robinson2010], borrowed from RNA-seq, computes library-specific
normalisation factors by trimming the distribution of log-fold changes between
each sample and a reference.

```{r tmm, eval = rlang::is_installed("edgeR")}
data_tmm <- tmm_pq(data_fungi_mini)
```

### `vst_pq()` — Variance Stabilising Transformation

VST [@love2014] fits a negative-binomial model to stabilise the
mean-variance relationship, making counts more suitable for methods that
assume homoscedastic data.

```{r vst, eval = rlang::is_installed("DESeq2")}
data_vst <- vst_pq(data_fungi_mini)
```

### `mcknight_residuals_pq()` — depth-robust alpha diversity

Instead of normalising the count table, this function computes residuals from
a regression of log-richness on log-depth and stores them in `sample_data`.
The residuals represent richness corrected for depth variation — a
depth-robust alpha diversity metric proposed by @mcknight2019 and used in
large-scale soil fungal surveys [@mikryukov2023].

```{r mcknight}
data_res <- mcknight_residuals_pq(data_fungi_mini)
head(sample_data(data_res)$mcknight_residuals)
```

---

## Comparing methods via ordination

A PCoA on Bray-Curtis dissimilarity illustrates how strongly the choice of
normalisation shapes the ordination landscape.

```{r ordination-comparison, fig.height=10, fig.cap="PCoA (Bray-Curtis) on raw counts, TSS, Hellinger, and log1p-transformed data. Colour encodes the 'Height' variable.", eval = rlang::is_installed("vegan")}
ord_raw <- phyloseq::ordinate(data_fungi_mini, method = "PCoA", distance = "bray")
ord_tss <- phyloseq::ordinate(data_tss, method = "PCoA", distance = "bray")
ord_hell <- phyloseq::ordinate(data_hell, method = "PCoA", distance = "bray")
ord_log1p <- phyloseq::ordinate(data_log1p, method = "PCoA", distance = "bray")

p_raw <- phyloseq::plot_ordination(
  data_fungi_mini, ord_raw, color = "Height"
) + ggplot2::ggtitle("Raw counts")

p_tss <- phyloseq::plot_ordination(
  data_tss, ord_tss, color = "Height"
) + ggplot2::ggtitle("TSS")

p_hell <- phyloseq::plot_ordination(
  data_hell, ord_hell, color = "Height"
) + ggplot2::ggtitle("Hellinger")

p_log1p <- phyloseq::plot_ordination(
  data_log1p, ord_log1p, color = "Height"
) + ggplot2::ggtitle("log1p")

patchwork::wrap_plots(p_raw, p_tss, p_hell, p_log1p, ncol = 2) +
  patchwork::plot_layout(guides = "collect")
```

---

## Model-level depth correction (no table transformation)

Some MiscMetabar functions correct for sequencing depth **without**
transforming the OTU table, by incorporating library size directly into the
statistical model.  These are complementary to — not replacements for — the
table-level normalisations above. If you transform the table, you should not use these model-level corrections (set correction_for_sample_size = FALSE).

### `adonis_pq()` — PERMANOVA with library-size covariate

When `correction_for_sample_size = TRUE`, `adonis_pq()` prepends
`sample_size` (i.e. `sample_sums(physeq)`) to the PERMANOVA formula, as
recommended by @weiss2017:

```
sample_size + Biological_Effect ~ distance_matrix
```

Library-size variation is partitioned out as the first term, so the
remaining variance attributed to the biological factor is not confounded by
depth.  The raw count table is used as-is; only the formula changes.

### `hill_pq()` / `hill_tuckey_pq()` — alpha diversity with sqrt-depth residuals

When `correction_for_sample_size = TRUE` (the default), these functions:

1. Compute Hill numbers from the **raw** OTU table.
2. Fit `lm(hill_values ~ sqrt(read_numbers))` per diversity order.
3. Use the **residuals** as the response in Tukey ANOVA.

The square-root of read count linearises the typical depth–richness
relationship; the residuals represent diversity corrected for depth without
discarding any data. Note that the Hill-number values **displayed** are
always computed from raw counts — only the *statistical test* uses residuals.

This is conceptually similar to `mcknight_residuals_pq()` but integrated
into the plotting workflow rather than stored as a new column in `sam_data`.

---

<!--
## Decision guide

```
Are you doing alpha diversity?
├── Yes → mcknight_residuals_pq()  (depth-robust, no data loss)
│         or rarefy_pq(n ≥ 10)     (classical, accounts for sampling effort)
└── No  →
    Are you testing differential abundance?
    ├── Yes → vst_pq() or css_pq() or tmm_pq()
    └── No  (ordination / clustering) →
        Is the data very zero-inflated?
        ├── Yes → gmpr_pq() or srs_pq() or transform_pq("rclr")
        └── No  → transform_pq("hellinger") or transform_pq("clr")
```

**General rules of thumb**

- Avoid rarefying if library sizes vary by more than ×10 and sample sizes are small; use GMPR or SRS instead.
- CLR / rCLR are theoretically sound for compositional data but sensitive to zero-handling; prefer Hellinger when sparsity is high.
- For model-based methods (DESeq2, edgeR) use VST or TMM rather than pre-normalising.
- McKnight residuals are specifically designed as a replacement for rarefaction in alpha diversity estimation.

---
-->

## References
