---
title: "Examples with published dataset: Tengeler"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examples with published dataset: Tengeler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This tutorial explore a phyloseq version of the dataset from Tengeler et al. (2020) available in the `mia` package.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.width = 10
)
```

# Load library 

```{r}
library("MicrobiotaProcess")
library("MiscMetabar")
library("ggplot2")
library("patchwork")
library("iNEXT")
```


```{r, eval=FALSE}
?Tengeler2020
```


# Import dataset in phyloseq format


```{r}
data(Tengeler2020_pq)
ten <- Tengeler2020_pq
summary_plot_pq(ten)
```

# Alpha-diversity analysis

```{r}
hill_pq(ten, "patient_status", one_plot = TRUE)
```

```{r}
res_inext <-
  iNEXT_pq(ten,
    datatype = "abundance",
    merge_sample_by = "patient_status_vs_cohort",
    nboot = 5
  )
ggiNEXT(res_inext)
```

```{r}
accu_plot(
  ten,
  fact = "sample_name",
  add_nb_seq = TRUE,
  by.fact = TRUE,
  step = 100
) + theme(legend.position = c(.8, .6))
```

# Explore taxonomy

```{r}
# library(metacoder)
# heat_tree_pq(
#   ten,
#   node_size = n_obs,
#   node_color = nb_sequences,
#   node_label = taxon_names,
#   tree_label = taxon_names,
#   node_size_trans = "log10 area"
# )
```

```{r}
treemap_pq(ten, lvl1 = "Order", lvl2 = "Family")
```

# Beta-diversity analysis : effect of patient status and cohort

```{r}
circle_pq(ten, "patient_status")
```

```{r, eval = requireNamespace("ComplexUpset", quietly = TRUE) && MiscMetabar::is_ggplot2_compatible()}
upset_pq(ten, "patient_status_vs_cohort")
```

```{r}
ggvenn_pq(clean_pq(ten, force_taxa_as_columns = TRUE),
  "cohort",
  rarefy_before_merging = TRUE
) +
  theme(legend.position = "none")
```


```{r}
ten_control <- clean_pq(subset_samples(ten, patient_status == "Control"))
# p_control <- heat_tree_pq(
#   ten_control,
#   node_size = n_obs,
#   node_color = nb_sequences,
#   node_label = taxon_names,
#   tree_label = taxon_names,
#   node_size_trans = "log10 area"
# )

ten_ADHD <- clean_pq(subset_samples(ten, patient_status == "ADHD"))
# p_ADHD <- heat_tree_pq(
#   ten_ADHD,
#   node_size = n_obs,
#   node_color = nb_sequences,
#   node_label = taxon_names,
#   tree_label = taxon_names,
#   node_size_trans = "log10 area"
# )
#
# p_control + ggtitle("Control") + p_ADHD + ggtitle("ADHD")
```


```{r}
knitr::kable(track_wkflow(list(
  "All samples" = ten,
  "Control samples" = ten_control,
  "ADHD samples" = ten_ADHD
)))
```

```{r}
adonis_pq(ten, "cohort + patient_status")
```


```{r}
ten@tax_table <- phyloseq::tax_table(cbind(
  ten@tax_table,
  "Species" = taxa_names(ten)
))

biplot_pq(subset_taxa_pq(ten, taxa_sums(ten) > 3000),
  merge_sample_by = "patient_status",
  fact = "patient_status",
  nudge_y = 0.4
)
```


```{r}
multitax_bar_pq(ten, "Phylum", "Class", "Order", "patient_status")
```


```{r}
multitax_bar_pq(ten, "Phylum", "Class", "Order", "patient_status",
  nb_seq = FALSE, log10trans = FALSE
)
```

# Differential abundance analysis

```{r}
plot_deseq2_pq(ten,
  contrast = c("patient_status", "ADHD", "Control"),
  taxolev = "Genus"
)
```

```{r}
LEfSe <- diff_analysis(
  ten,
  classgroup = "patient_status",
  mlfun = "lda",
  ldascore = 2,
  p.adjust.methods = "bh"
)
library(ggplot2)
ggeffectsize(LEfSe) +
  scale_color_manual(values = c(
    "#00AED7",
    "#FD9347"
  )) +
  theme_bw()
```





# Session information

```{r}
sessionInfo()
```

# References


Tengeler, A.C., Dam, S.A., Wiesmann, M. et al. Gut microbiota from persons with attention-deficit/hyperactivity disorder affects the brain in mice. Microbiome 8, 44 (2020). https://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-020-00816-x


