---
title: "Function timing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Function timing}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

```{r setup, message = FALSE}
library(MiscMetabar)
```

## Why this article exists

Metabarcoding datasets vary by two to three orders of magnitude in size — a
small pilot may have a few dozen samples and a few hundred OTUs, while a
regional-scale dataset can reach thousands of samples and tens of thousands
of OTUs. Functions in `MiscMetabar` were written with the small / medium case
in mind, and several of them have a cost that grows non-linearly with the
input.

This article reports wall-clock timings of the main exported functions on two
reference datasets shipped with the package, so you can:

- pick reasonable defaults for permutation- and bootstrap-based analyses,
- decide which calls are worth wrapping in `targets::tar_target()` to avoid
  recomputing them,
- and know which functions to subset (samples or taxa) before running them on
  a larger dataset.

The two reference datasets are:

| Dataset | Samples | Taxa | Use |
|---|---|---|---|
| `data_fungi_mini` | 45 | 137 | Quick sanity checks; examples; unit tests |
| `data_fungi` | 185 | 1420 | Realistic medium-scale dataset; what most users will plug in |

The timings below were collected on the package's CI machine
(`x86_64-pc-linux-gnu`, 4 cores, 16 GB RAM) with R 4.5.2 and one replicate
per call. Numbers should be read as orders of magnitude, not as benchmarks.

## Timing table

```{r table, echo = FALSE}
csv_path <- system.file(
  "benchmark", "function_timings.csv",
  package = "MiscMetabar"
)
if (nzchar(csv_path) && file.exists(csv_path)) {
  tab <- utils::read.csv(csv_path, stringsAsFactors = FALSE)
  wide <- stats::reshape(
    tab,
    direction = "wide",
    idvar = "fonction",
    timevar = "dataset"
  )
  colnames(wide) <- sub("^elapsed_s\\.", "", colnames(wide))
  knitr::kable(
    wide,
    col.names = c("Function", "data_fungi_mini (s)", "data_fungi (s)"),
    align = c("l", "r", "r"),
    caption = "Elapsed time (seconds) per function and dataset. One replicate."
  )
} else {
  message(
    "Timing CSV not found. Run `Rscript inst/benchmark/function_timings.R` ",
    "from the package root to regenerate it."
  )
}
```

## How to speed up your own analysis

### Subset before computing

For every function that ends in `_pq`, the cost is dominated either by the
number of samples (permutation tests, ordinations) or by the number of taxa
(taxonomy checks, plotting). Subset first:

```{r, eval = FALSE}
# Keep only the most abundant taxa
data_fungi_subset <- subset_taxa_pq(data_fungi, taxa_sums(data_fungi) > 1000)

# Or merge samples by a grouping variable before plotting
data_fungi_merged <- merge_samples2(data_fungi, "Height")
```

### Cache permutation-based results

`adonis_rarperm_pq()`, `hill_test_rarperm_pq()`, `var_par_rarperm_pq()`, and
`hill_acc_pq(type = "sample")` are the most expensive functions because they
run a full permutation loop. They are also the most worth caching with
`targets` or `memoise`:

```{r, eval = FALSE}
library(memoise)
adonis_pq_m <- memoise(adonis_pq)
adonis_pq_m(data_fungi, "Height") # slow the first time
adonis_pq_m(data_fungi, "Height") # instant the second time
```

### Lower iteration counts for exploratory work

Most permutation-based functions take an iteration argument (`nperm`,
`n_permutations`, `nboot`, `n_simulations`). The defaults are tuned for
final, publishable analyses. For exploratory work, lowering them by an order
of magnitude is almost always fine:

```{r, eval = FALSE}
# Default: 999 permutations
adonis_pq(data_fungi_mini, "Height")

# Exploration: 99 permutations
adonis_pq(data_fungi_mini, "Height", nperm = 99)
```

The function defaults stay at the published values for reproducibility; the
choice is yours to lower them in your own code.

### Use `divent::` and `iNEXT.4steps::` for very large datasets

For datasets larger than `data_fungi`, the underlying packages
`divent` (Hill numbers) and `iNEXT.4steps` (coverage-based rarefaction)
expose lower-level functions that skip phyloseq slot lookups and can be
called directly on the OTU matrix.

## Reproducing this table

The script that generates `inst/benchmark/function_timings.csv` is shipped
with the package:

```{r, eval = FALSE}
system.file("benchmark", "function_timings.R", package = "MiscMetabar")
```

Run it from the package source root:

```bash
Rscript inst/benchmark/function_timings.R
```

Re-rendering this article will then pick up the updated CSV.

## Internal speedups since 0.16.5

Recent versions of `MiscMetabar` include internal rewrites of the most
expensive helpers. The user-facing behaviour is unchanged; only the wall
time differs. Where the speedup is documented (`NEWS.md`), it is the
result of vectorising one or more inner loops or factoring out repeated
rarefaction work, **not** of changing the scientific defaults. See
`NEWS.md` for the version-by-version list.
